---
title: "RMS Titanic"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Submission
<!-- ------------------------- -->

Make sure to commit both the challenge report (`report.md` file) and supporting files (`report_files/` folder) when you are done! Then submit a link to Canvas. **Your Challenge submission is not complete without all files uploaded to GitHub.**


```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- (List all variables here)
Class, Sex, Age, Survived, n


### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Write your observations here
- Are there any differences?
  - (Your response here)
  Yes the summarized number is different but not by a signifigant amount. 
- If yes, what might account for those differences?
  - (Your response here)
  The wikipedia article warns its souls count is an estimate so perhaps the data set provided in this assignment and the article use 2 slightly different estimations of titanic's soul on board count.
  - ...

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
library(ggplot2)
library(tidyverse)
library(dplyr)
df_titanic <- as_tibble(Titanic)
df_survivors <- df_titanic %>%
  group_by(Class,Sex) %>%
  filter(Survived=="Yes") %>%
  summarise(tt= sum(n))

ggplot(df_survivors, aes(x=Class,y=tt,fill=Sex,))+
  geom_bar(stat = "identity",position = "dodge")+
  ggtitle("Count of Survivors by Class and Sex")+
  xlab("Passenger Class")+
  ylab("Count of Survivors")
  theme_dark()
 
```

**Observations**:

took super long to figure out but i got it eventually
# Deeper Look
I noticed in general females survived more then men. Also the high proportion of male crew survivors to female ones is probably due to sexist attitudes of the time period leading to a mostly male crew. 
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
library(ggplot2)
library(tidyverse)
library(dplyr)
df_prop <-
  df_titanic %>%
  group_by(Class, Sex) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop

ggplot(df_prop, aes(x=Class,y=Prop,fill=Sex,))+
  geom_bar(stat = "identity",position = "dodge")+
  ggtitle("Count of Survivors by Class and Sex")+
  xlab("Passenger Class")+
  ylab("Count of Survivors")
  theme_dark()
 
```

**Observations**:

- Write your observations here.
The code worked as expected
- Is there anything *fishy* going on in your plot?
  - (Your response here)
  Yeah the overall passenger and crew survival rates are in the 30ish percent zone but all the bars are way way above that so thats pretty weird. 

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
library(ggplot2)
library(tidyverse)
library(dplyr)
df_prop <-
  df_titanic %>%
  group_by(Class, Sex,Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop

ggplot(df_prop, aes(x=Class,y=Prop,fill=Sex))+
  geom_bar(stat = "identity",position = "dodge")+
  facet_wrap(~Age)+
  ggtitle("Count of Survivors by Class and Sex")+
  xlab("Passenger Class")+
  ylab("Count of Survivors")
  theme_dark()
 
```

**Observations**:

- (Write your observations here.)
It worked but took some effort it divided them into 2 plots from q4 one for children one for adults.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - (Your response here)
  Its fishy because the dataset has a few of the children data values messed up such that the proportion shows all children in said category survived. This wasnt plainly obvious in the first graph but it was enough to look fishy.

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
